### 2017-11-07 submission/manuscript.Rmd
discussion edits

Our study identifies clear diversity changes both at the community level and within individual genera that are present in indivdiuals with carcinoma versus those without the disease. Although there was a step wise decrease in diversity from control to adenoma to carcinoma; this did not translate into large effect sizes for the relative risk of either of these two conditions. Even though CRC associated genera increases the relative risk of carcinoma it does not increase the relative risk of adenoma. This information suggests that these specific genera may not be the primary members of the microbial community that contributes to the formation of an adenoma but is for carcinoma. Additionally, our data shows that by using the whole community our models perform better then when they only use the CRC associated genera. CRC associated genera are clearly important to carcinoma but the context or community in which these microbes are a part of can drastically increase the ability of models to make predictions. This data supports the concept that small localized changes within the community may be occuring that are important in the disease progression of colorectal cancer and that they may not directly involve CRC associated genera.

The driver-passenger model of the microbial role in CRC, as summarized by Flynn [@flynn_metabolic_2016], can be supported with this data for carcinoma but not necessarily for adenoma. The drasitically increased relative risk of disease when considering the CRC associated genera is highly supportive of this type of process, especially in the context of increasing relative risk with more CRC associated genera positivity. It is also possible that in a driver-passenger scenario it is possible that simply having the driver present or only identifying the passenger is a good enough proxy that the event is occuring. This would account for the observation that there is no constant additive effect on relative risk for increasing positivity. Additionally, the initial establishment of the driver within the system is also dependent on the community that is present and this is supported by the observation that when adding the community context to our models along with the CRC associated genera the model AUC increases.

Our carcinoma observations fit the driver-passenger model and support this concept within the framework of the transition from adenoma to carcinoma. In contrast, with the present data we can only suggest that the adenoma observations might fit with this model but the changes that occur at this timepoint are small and possibly focal to the adenoma or specific location. The stepwise decrease in diversity suggests that the adenoma community is not normal but this change is subtle. Although there may be localized changes that do depend on the driver-passenger model, supported by the one, two, and four positive CRC associated genera in tissue relative risk increases [Figure 3C], there may be another process involved that ultimately exacerbates the condition from a subtle localized change to a global community one. The poor performance of the Random Forest models for classifying adenoma based only on the microbiota would suggest that this is the case. It is possible to hypothesize that at early stages of the diease, how the host interacts to these subtle changes could be the catlyst that ultimately leads to this larger global dysfunctional community.

@@

By aggregating together a large collection of studies from both feces and tissue we are able to provide information in support of the driver-passenger model in the context of carcinoma. However, within the context of adenoma it is less clear that this relationship exists. These observations highlight the importance of power and sample number considerations when considering investigations into the microbiota and adenoma due to the subtle changes in the community.

>>>>

Our study identifies clear diversity changes both at the community level and within individual genera that are present in individuals with carcinoma versus those without the disease. Although there was a step wise decrease in diversity from control to adenoma to carcinoma; this did not translate into large effect sizes for the relative risk of either of these two conditions. Even though CRC associated genera increases the relative risk of carcinoma it does not consistently increase the relative risk of adenoma. This information suggests that these specific genera are important in carcinoma genesis but may not be the primary members of the microbial community that contribute to the formation of an adenoma. Additionally, our data shows that by using the whole community our models perform better then when they only use the CRC associated genera. CRC associated genera are clearly important to carcinoma but the context or community in which these microbes are a part of can drastically increase the ability of models to make predictions. This data supports the concept that small localized changes within the community may be occuring that are important in the disease progression of colorectal cancer and that they may not directly involve CRC associated genera.

The driver-passenger model of the microbial role in CRC, as summarized by Flynn [@flynn_metabolic_2016], can be supported with this data for carcinoma but not necessarily for adenoma. The drasitically increased relative risk of disease, when considering the CRC associated genera, is highly supportive of this type of process, especially in the context of increasing relative risk with more CRC associated genera positivity. It is also possible that in a driver-passenger scenario that simply having the driver present or only identifying the passenger is a good enough proxy that the event is occuring. This would account for the observation that there is no constant additive effect on relative risk for increasing positivity. Additionally, the initial establishment of the driver within the system is also dependent on the community that is present and this is supported by the observation that when adding the community context to our models along with the CRC associated genera the model AUC increases.

Our carcinoma observations fit the driver-passenger model and support this concept within the framework of the transition from adenoma to carcinoma. In contrast, with the present data we can only suggest that the adenoma observations might fit with this model but the changes that occur at this timepoint are small and possibly focal to the adenoma. The stepwise decrease in diversity suggests that the adenoma community is not normal but this change is subtle. Although there may be localized changes that do depend on the driver-passenger model, supported by an increased relative risk for one, two, and four positive CRC associated genera in tissue [Figure 3C], there may be other processes involved that ultimately exacerbates the condition from a subtle localized change to a global community one. The poor performance of the Random Forest models for classifying adenoma based only on the microbiota would suggest that this is the case. It is possible to hypothesize that at early stages of the diease, how the host interacts to these subtle changes could be the catlyst that ultimately leads to this larger global dysfunctional community.

@@

By aggregating together a large collection of studies from both feces and tissue we are able to provide information in support of the importance of the bacterial community in both adenoma and carcinoma. We are also able to provide support for the driver-passenger model in the context of carcinoma. However, within the context of adenoma it is less clear that this relationship exists. These observations highlight the importance of power and sample number considerations when considering investigations into the microbiota and adenoma due to the subtle changes in the community.

### 2017-11-07 submission/manuscript.tex
discussion edits

Our study identifies clear diversity changes both at the community level
and within individual genera that are present in indivdiuals with
carcinoma versus those without the disease. Although there was a step
wise decrease in diversity from control to adenoma to carcinoma; this
did not translate into large effect sizes for the relative risk of
either of these two conditions. Even though CRC associated genera
increases the relative risk of carcinoma it does not increase the
relative risk of adenoma. This information suggests that these specific
genera may not be the primary members of the microbial community that
contributes to the formation of an adenoma but is for carcinoma.
Additionally, our data shows that by using the whole community our
models perform better then when they only use the CRC associated genera.
CRC associated genera are clearly important to carcinoma but the context
or community in which these microbes are a part of can drastically
increase the ability of models to make predictions. This data supports
the concept that small localized changes within the community may be
occuring that are important in the disease progression of colorectal
cancer and that they may not directly involve CRC associated genera.

The driver-passenger model of the microbial role in CRC, as summarized
by Flynn {[}2{]}, can be supported with this data for carcinoma but not
necessarily for adenoma. The drasitically increased relative risk of
disease when considering the CRC associated genera is highly supportive
of this type of process, especially in the context of increasing
relative risk with more CRC associated genera positivity. It is also
possible that in a driver-passenger scenario it is possible that simply
having the driver present or only identifying the passenger is a good
enough proxy that the event is occuring. This would account for the
observation that there is no constant additive effect on relative risk
for increasing positivity. Additionally, the initial establishment of
the driver within the system is also dependent on the community that is
present and this is supported by the observation that when adding the
community context to our models along with the CRC associated genera the
model AUC increases.

Our carcinoma observations fit the driver-passenger model and support
this concept within the framework of the transition from adenoma to
carcinoma. In contrast, with the present data we can only suggest that
the adenoma observations might fit with this model but the changes that
occur at this timepoint are small and possibly focal to the adenoma or
specific location. The stepwise decrease in diversity suggests that the
adenoma community is not normal but this change is subtle. Although
there may be localized changes that do depend on the driver-passenger
model, supported by the one, two, and four positive CRC associated
genera in tissue relative risk increases {[}Figure 3C{]}, there may be
another process involved that ultimately exacerbates the condition from
a subtle localized change to a global community one. The poor
performance of the Random Forest models for classifying adenoma based
only on the microbiota would suggest that this is the case. It is
possible to hypothesize that at early stages of the diease, how the host
interacts to these subtle changes could be the catlyst that ultimately
leads to this larger global dysfunctional community.

>>>>

Our study identifies clear diversity changes both at the community level
and within individual genera that are present in individuals with
carcinoma versus those without the disease. Although there was a step
wise decrease in diversity from control to adenoma to carcinoma; this
did not translate into large effect sizes for the relative risk of
either of these two conditions. Even though CRC associated genera
increases the relative risk of carcinoma it does not consistently
increase the relative risk of adenoma. This information suggests that
these specific genera are important in carcinoma genesis but may not be
the primary members of the microbial community that contribute to the
formation of an adenoma. Additionally, our data shows that by using the
whole community our models perform better then when they only use the
CRC associated genera. CRC associated genera are clearly important to
carcinoma but the context or community in which these microbes are a
part of can drastically increase the ability of models to make
predictions. This data supports the concept that small localized changes
within the community may be occuring that are important in the disease
progression of colorectal cancer and that they may not directly involve
CRC associated genera.

The driver-passenger model of the microbial role in CRC, as summarized
by Flynn {[}2{]}, can be supported with this data for carcinoma but not
necessarily for adenoma. The drasitically increased relative risk of
disease, when considering the CRC associated genera, is highly
supportive of this type of process, especially in the context of
increasing relative risk with more CRC associated genera positivity. It
is also possible that in a driver-passenger scenario that simply having
the driver present or only identifying the passenger is a good enough
proxy that the event is occuring. This would account for the observation
that there is no constant additive effect on relative risk for
increasing positivity. Additionally, the initial establishment of the
driver within the system is also dependent on the community that is
present and this is supported by the observation that when adding the
community context to our models along with the CRC associated genera the
model AUC increases.

Our carcinoma observations fit the driver-passenger model and support
this concept within the framework of the transition from adenoma to
carcinoma. In contrast, with the present data we can only suggest that
the adenoma observations might fit with this model but the changes that
occur at this timepoint are small and possibly focal to the adenoma. The
stepwise decrease in diversity suggests that the adenoma community is
not normal but this change is subtle. Although there may be localized
changes that do depend on the driver-passenger model, supported by an
increased relative risk for one, two, and four positive CRC associated
genera in tissue {[}Figure 3C{]}, there may be other processes involved
that ultimately exacerbates the condition from a subtle localized change
to a global community one. The poor performance of the Random Forest
models for classifying adenoma based only on the microbiota would
suggest that this is the case. It is possible to hypothesize that at
early stages of the diease, how the host interacts to these subtle
changes could be the catlyst that ultimately leads to this larger global
dysfunctional community.
